Biogeography of crop progenitors and wild plant resources during the transition to agriculture in West Asia

### 2024-11-28 analysis/analysis.qmd
Update predictor datasets

```{r dependencies}
library("cowplot")
library("dplyr")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("rpaleoclim")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

w_asia <- st_bbox(c(xmin = 30, xmax = 50, ymin = 25, ymax = 40), crs = 4326)
s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
```

@@

neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")
```

@@

```{r data-climate}
raw_data("paleoclim") |>
  dir_ls() |>
  map(dir_ls, glob = "*.tif") |>
  map(read_stars) |>
  map(\(x) set_names(x, path_ext_remove(path_file(names(x))))) ->
  climate

# Tidy up names
names(climate) |>
  basename() |>
  str_remove(regex("chelsa_", ignore_case = TRUE)) |>
  str_split_i("_", 1) ->
  names(climate)
```

```{r training-climate}
flora |> 
  mutate(
    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
    # TODO: where is the extra geometry column coming from?
    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
  ) ->
  flora
```

>>>>

```{r dependencies}
library("cowplot")
library("dplyr")
library("dotenv")
library("forcats")
library("fs")
library("furrr")
library("ggplot2")
library("ggrepel")
library("ggspatial")
library("gt")
library("here")
library("khroma")
library("progressr")
library("purrr")
library("ranger")
library("readr")
library("recipes")
library("rgbif")
library("scales")
library("sf")
library("stars")
library("stringr")
library("tibble")
library("tidymodels")
library("tidyr")

@@

w_asia <- st_bbox(c(xmin = 25, xmax = 55, ymin = 22.5, ymax = 42.5), crs = 4326)
s_levant <- st_bbox(c(xmin = 34, xmax = 39, ymin = 29, ymax = 34), crs = 4326)
```

@@

neolithic <- c("PPNA", "EPPNB", "MPPNB", "LPPNB/C", "Pottery Neolithic")

climate_periods <- fct_inorder(c(
  lgm = "LGM (21 ka)",
  hs1 = "HS1 (17.0-14.7 ka)",
  ba  = "BA (14.7-12.9 ka)",
  yds = "YDS (12.9-11.7 ka)",
  eh  = "EH (11.7-8.3 ka)",
  cur = "current"
))
```

@@

Bioclimatic data – the standard, and what allows us to do hindcasting.

Topographic data derived from a DEM:

* Aspect and slope – proxies for solar radiation, but only if high resolution [@AustinVanNiel2011; @LeempoelEtAl2015]
* TWI – proxy for soil moisture, important for arid environments [@KopeckyCizkova2010; @CamposEtAl2016; DiVirgilioEtAl2018]

Soil data – important for plants [@DubuisEtAl2013; @ModEtAl2016; @VelazcoEtAl2017] but not all variables from SoilGrids are reliable: so we just use soil texture (clay, silt, sand) and pH at the surface (0-5 cm depth) [@MillerEtAl2024].

```{r data-terrain}
# Reacquire derived_data/srtm_15plus_w_asia.tif if needed:
if (!file_exists(derived_data("dem", "srtm_15plus_w_asia.tif"))) {
  message("Redownloading DEM data...")
  source(here("data", "data-dem.R"))
}

dem <- read_stars(derived_data("dem", "srtm_15plus_w_asia.tif"))
names(dem) <- "elevation"

# TODO: slope
# TODO: aspect
# TODO: twi
```

```{r data-soil}
# Reacquire derived_data/soilgrids if needed:
if (length(dir_ls(derived_data("soilgrids"), glob = "*.tif")) < 1) {
  message("Redownloading SoilGrids data...")
  source(here("data", "data-soilgrids.R"))
}

soil <- map(dir_ls(derived_data("soilgrids")), read_stars)
names(soil) <- str_split_i(basename(names(soil)), "_", 2)
soil <- do.call(c, soil)
```

```{r data-climate}
# Reacquire derived_data/paleoclim if needed:
if (length(dir_ls(derived_data("paleoclim"), glob = "*.tif")) < 1) {
  message("Redownloading Paleoclim data...")
  source(here("data", "data-paleoclim.R"))
}

climate <- read_stars(
  derived_data(
    "paleoclim", 
    paste0("paleoclim_", names(climate_periods), "_2_5m_w_asia.tif")
  ),
  along = NA
)

names(climate) <- names(climate_periods)

climate <- st_redimension(climate, name = "period") |>
  split("band")
```

```{r data-environment}
# TODO: resample to common res
# TODO: reproject
# TODO: combine
# TODO: fill NAs?
```

```{r extract-predictor-data}
flora |> 
  mutate(
    occ = map(occ, \(x) cbind(x, st_extract(climate$cur, gbif_to_sf(x)))),
    # TODO: where is the extra geometry column coming from?
    occ = map(occ, \(x) as_tibble(select(x, -geometry))),
    # TODO: can this happen earlier?
    occ = map(occ, \(x) mutate(x, present = factor(present, c(FALSE, TRUE))))
  ) ->
  flora
```
